---
title: "COVID 19 - Analysis of a Pandemic"

date: "2024-04-24"
output: pdf_document
---

## Description

**COVID-19 Cases and Deaths by Political Boundaries** -
The COVID-19 pandemic was the first truly global pandemic of our lifetime. How it
affected and interacted with populations varied widely from country to country. 
Treatment protocols and detection methods also differed widely. Some of the most 
affected nations also had some of the best treatment protocols while some 
countries may have given a false impression of low numbers simply due to poor 
detection methods.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(lubridate)
library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## R Markdown

This information regarding COVID-19 is provided from Johns Hopkins University and 
is hosted on git hub. (https://github.com/CSSEGISandData/COVID-19) 

```{r download}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])

opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Transforming the Data Set for Comparison
```{r data transform, include=TRUE}
global_cases_pivot <- global_cases %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date" , values_to = "cases") %>% select(-c(Lat,Long))

global_deaths_pivot <- global_deaths %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date" , values_to = "deaths") %>% select(-c(Lat,Long))


global_all <- global_cases_pivot %>% full_join(global_deaths_pivot) %>% rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>% mutate(date = mdy(date))

global_all <- global_all %>% unite("Combined_Key", c(Province_State, Country_Region), sep=", ", na.rm = TRUE, remove = FALSE)


uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))


global_all_uid <- global_all %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

global_by_country <- global_all_uid %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()


global_totals <- global_by_country %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))


opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Analyzing the Data Set to Choose Two Countries
```{r data visualizations, include=TRUE}

global_totals_per_thou <- global_totals %>% group_by(Country_Region) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases > 0, population > 0)


global_totals_per_thou %>% filter(deaths_per_thou > 0) %>% slice_min(deaths_per_thou, n=15)
global_totals_per_thou %>% slice_max(deaths_per_thou, n=15)




opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```
## Comparing Country with Low Deaths to Country with High Deaths
Based off of the results of the 15 lowest death rates versus the 15 highest 
death rates, Peru and Nigeria were chosen for the one to one comparison. This 
analysis will also show the differences in the predictive modeling for both 
data sets.

```{r data graph, include=TRUE}



low_country <- "Nigeria"
low_country_numbers <- global_all_uid %>% filter(Country_Region == low_country)

low_country_numbers <- low_country_numbers %>% mutate(deaths_per_thou = deaths * 1000 / Population) %>%
  mutate(cases_per_thou = cases * 1000/Population) %>% 
  select(date, cases, cases_per_thou, deaths, deaths_per_thou, Population) %>% ungroup()

mod_low <- lm(deaths_per_thou ~ cases_per_thou, data = low_country_numbers)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
low_country_numbers %>% mutate(pred = predict(mod_low))
country_numbers_low_w_pred <- low_country_numbers %>% mutate(pred = predict(mod_low))


high_country <- "Peru"

high_country_numbers <- global_all_uid %>% filter(Country_Region == high_country)

high_country_numbers <- high_country_numbers %>% mutate(deaths_per_thou = deaths * 1000 / Population) %>%
  mutate(cases_per_thou = cases * 1000/Population) %>% 
  select(date, cases, cases_per_thou, deaths, deaths_per_thou, Population) %>% ungroup()

mod_high <- lm(deaths_per_thou ~ cases_per_thou, data = high_country_numbers)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
high_country_numbers %>% mutate(pred = predict(mod_high))
country_numbers_high_w_pred <- high_country_numbers %>% mutate(pred = predict(mod_high))


low_country_numbers %>% ggplot(aes(x=date,y=cases_per_thou)) + geom_line(aes(color="cases_per_thou")) + geom_point(aes(color="cases_per_thou")) + geom_line(aes(y=deaths_per_thou, color="deaths_per_thou")) + geom_point(aes(y=deaths_per_thou, color = "deaths_per_thou")) + scale_y_log10() + theme(legend.position="bottom",axis.text.x=element_text(angle=90)) + labs(title = str_c("COVID-19 in ", low_country), y=NULL)

high_country_numbers %>% ggplot(aes(x=date,y=cases_per_thou)) + geom_line(aes(color="cases_per_thou")) + geom_point(aes(color="cases_per_thou")) + geom_line(aes(y=deaths_per_thou, color="deaths_per_thou")) + geom_point(aes(y=deaths_per_thou, color = "deaths_per_thou")) + scale_y_log10() + theme(legend.position="bottom",axis.text.x=element_text(angle=90)) + labs(title = str_c("COVID-19 in ", high_country), y=NULL)


country_numbers_low_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue") + geom_point(aes(x = cases_per_thou, y = pred),color="red") + ylab("Deaths per Thousand") + xlab("Cases per Thousand") + ggtitle("COVID-19 Cases and Deaths for Nigeria", subtitle = "Blue = Actual vs. Red = Predicted") + theme(legend.position = "bottom")

country_numbers_high_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue") + geom_point(aes(x = cases_per_thou, y = pred),color="red") + ylab("Deaths per Thousand") + xlab("Cases per Thousand") + ggtitle("COVID-19 Cases and Deaths for Peru", subtitle = "Blue = Actual vs. Red = Predicted")




##print(bar_graph_year)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Bias Writeup

**Possible Bias in this Analysis** -
There is not enough known about the differences in treating and detecting
COVID-19 in Peru versus Nigeria. The visual and numerical information cannot tell 
the entire story of how one country may have successfully treated COVID-19 versus 
another. More research could be done to understand treatment and detection to 
allow for two very similar countries to be chosen for this analysis.


## Add Session Info
```{r data, include=TRUE}
sessionInfo()
```