---
title: "New York Shooting Deaths - A Time Analysis"

date: "2024-04-15"
output: pdf_document
---

## Description

**A Simple Demonstration of the Relationship Between Temperature and Homicide** -
Most people that have taken a psychology or statistics class will have 
heard the anecdote of ice cream sales rising and falling with homicide rates. 
This story is told as a cautionary tale of how correlation does not always equal 
causation. The ice cream sales and homicide rates actually rise and fall due to 
the temperature outside in the tale. When examining the data set, it was apparent 
that this relationship could be easily shown if true. Will shooting deaths rise 
and fall with the local recorded temperature?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## R Markdown

This information is provided from the City of New York Open Data website (https://data.cityofnewyork.us) regarding shooting death incidents.

```{r download}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
ny_shooting_deaths_data <- read_csv(url)
summary(ny_shooting_deaths_data)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Transforming the Data by Year
```{r data transform, include=TRUE}
ny_shooting_deaths_data <- ny_shooting_deaths_data %>% 
  mutate(OCCUR_DATE=as.Date(OCCUR_DATE, format = "%m/%d/%Y"))

ny_shooting_deaths_by_year <- ny_shooting_deaths_data %>% 
  group_by(incident_year = lubridate::floor_date(OCCUR_DATE, 'year')) %>% 
  summarize(n())

names(ny_shooting_deaths_by_year)[names(ny_shooting_deaths_by_year)=="n()"] <- "number_of_deaths"


ny_shooting_deaths_by_year %>% mutate(incident_year = year(incident_year))

opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```
## Plotting the Number of Shootings by Year
```{r data graph, include=TRUE}
bar_graph_year <- ny_shooting_deaths_by_year %>% ggplot(aes(x=incident_year,y=number_of_deaths)) + 
geom_bar(stat="sum",color="blue",width=.3, position=position_dodge(width=.6), show.legend=FALSE) +
  scale_x_date(date_breaks = "1 year", date_labels =  "%Y", guide = guide_axis(angle=45)) + 
  ylab("Total Shooting Deaths") + xlab("Year") + ggtitle("City of New York Shooting Deaths", subtitle = "Total by Year for 17 Years")

print(bar_graph_year)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Transforming the Data by Month
```{r data group by month, include=TRUE}
ny_shooting_deaths_by_month <- ny_shooting_deaths_data %>% 
  group_by(incident_month = lubridate::floor_date(OCCUR_DATE, 'month')) %>% 
  summarize(n())

names(ny_shooting_deaths_by_month)[names(ny_shooting_deaths_by_month)=="n()"] <- "number_of_deaths"

ny_shooting_deaths_avg_by_month <- aggregate(number_of_deaths ~ 
month(incident_month, label=TRUE),ny_shooting_deaths_by_month, mean)

names(ny_shooting_deaths_avg_by_month)[names(ny_shooting_deaths_avg_by_month)=="month(incident_month, label = TRUE)"] <- "incident_month"

opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```
## Plotting the Data by Month
```{r data graph month, include=TRUE}
bar_graph_month <- ny_shooting_deaths_avg_by_month %>% ggplot(aes(x=incident_month,y=number_of_deaths)) + 
     geom_bar(stat="identity",color="green",width=.3, position=position_dodge(width=.6), show.legend=FALSE) + 
     ylab("Average Number of Shooting Deaths") + xlab("Month") + geom_text(aes(label=round(number_of_deaths,0), vjust=0)) + ggtitle("City of New York Shooting Deaths", subtitle = "Average by Month for 17 Years")
 
print(bar_graph_month)

opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Linear Model for Temperature vs Shooting Deaths

The weather data in this section also came from the City of New York Open Data
Website. The data set is demonstrating the current baseline temperatures by month
for New York City while also predicting future temperatures due to climate change.
By extracting the baseline temperatures, one can use this data set to understand 
the relationship between shooting deaths and average temperature.
```{r linear model, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
weather_url = "https://data.cityofnewyork.us/api/views/hmdk-eidg/rows.csv?accessType=DOWNLOAD"

ny_weather_data <- read_csv(weather_url)

ny_weather <- ny_weather_data[ny_weather_data$Year == 'Baseline',]

ny_weather_temps <- ny_weather[ny_weather$`Month/Period`!= 'Annual' & 
ny_weather$`Month/Period`!= 'Winter' & 
  ny_weather$`Month/Period`!= 'Spring' & 
  ny_weather$`Month/Period`!= 'Summer' & ny_weather$`Month/Period`!= 'Fall',]

names(ny_weather_temps)[names(ny_weather_temps)=="Month/Period"] <- "Month"

names(ny_weather_temps)[names(ny_weather_temps)=="Mean Temperature"] <- "Mean_Temp"


ny_weather_temps <- ny_weather_temps %>% mutate(Mean_Temp = as.numeric(substr(Mean_Temp,1,4)))


ny_weather_temps <- ny_weather_temps %>% mutate(Month=substring(Month,1,3))

ny_weather_temps <- tibble::rowid_to_column(ny_weather_temps, "id")
ny_shooting_deaths_avg_by_month <- tibble::rowid_to_column(ny_shooting_deaths_avg_by_month, "id")

ny_weather_and_shooting_deaths <- merge(ny_shooting_deaths_avg_by_month, ny_weather_temps)


mod <- lm(number_of_deaths ~ Mean_Temp, data=ny_weather_and_shooting_deaths)

number_of_deaths_w_pred <- ny_weather_and_shooting_deaths %>% mutate(pred = predict(mod))

number_of_deaths_w_pred %>% ggplot() + geom_point(aes(x = Mean_Temp, 
                              y= number_of_deaths, color="Actual")) +  
geom_point(aes(x = Mean_Temp, y= pred, color="Predicted")) + 
  ylab("Number of Deaths") + 
  xlab("Average Daily Temperature in Farenheit") +
  ggtitle("Predicting Number of Shootings by Average Temperature", 
        subtitle = "Comparing Monthly Averages")


scatter.smooth(ny_weather_temps$Mean_Temp, 
               ny_shooting_deaths_avg_by_month$number_of_deaths,
               main='Shooting Deaths and Higher Temperatures',
               xlab = "Average Temperature by Month in Fahrenheit", 
ylab = "Average Shooting Deaths by Month for 17 Years")

opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Bias Writeup

**Possible Bias Present in Visual Plot** -
The total population at the time of the incident is not included in the number of 
shooting deaths data set. This information is better reported on a population 
percentage basis rather than just a total. If the change in the number of shooting 
deaths correlates with a change in population size, the increase or decrease in 
shooting deaths becomes less substantial.

**Possible Bias in Original Data Set** -
The original data set includes races of the victim and the perpetrator, but does 
not include an income bracket or a census block group description. Violent crimes 
often disproportionately affect people in lower income areas. Depending on where 
the incident occurred geographically, these areas may have a higher percentage of 
a specific race. A Borough is likely not the best delineation for the location 
of a violent incident as it is nonspecific. The latitude and longitude coordinates 
provided in the data set could be used to extrapolate census block group data. 
This could allow for a more thorough understanding of problems in that part of 
New York City that exacerbate violence. 


## Add Session Info
```{r data, include=TRUE}
sessionInfo()
```